\bf Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-16 manuscript/main.tex
remove old reg eigen proposition

	Given a twice-differentiable basis $\{b_m(t)\}$ of $L^2([0,1])$,  the $j$th functional predictor, regression coefficient, and PLS direction (suppressing iteration indices) are represented as
	\begin{equation*}
		X_{ij}(t) = \sum_{m=1}^\infty \theta_{ijm} \, b_m(t),\quad
		\beta_j(t) = \sum_{m=1}^\infty \eta_{jm} \, b_m(t),\quad
		%%%%%%%%%%%%%%%%%%%%%%
		\xi_j(t)
		=
		\sum_{m=1}^\infty
		\gamma_{jm}
		b_m(t)
	\end{equation*}	
	In practice, the full set of coefficients can not be obtained  with finite sample size, as functional data are measured on a finite grid. Thus, we truncate the expansion at $M$ terms and define the truncated basis space
	$
	\mathcal{B} := \operatorname{span}\bigl(b_1(t), \ldots, b_M(t)\bigr).
	$
	We choose a moderately large $M$ (e.g., 15 or 20) to capture functional variation without fine-tuning, as smoothness is handled via penalization (see Section~\ref{sec: Regularization of PLS Components}).
	The truncated expansions of the predictor and coefficient are denoted as
	\begin{equation*}
		\widetilde{X}_{ij}(t) := \sum_{m=1}^M \theta_{ijm} \, b_m(t), \quad
		\tilde{\beta}_j(t) := \sum_{m=1}^M \eta_{jm} \, b_m(t),
	\end{equation*}
	and our suggest method restricts each PLS direction to admit the following expansion:
	\begin{equation*}
		\xi_j(t) = \sum_{m=1}^M \gamma_{jm} \, b_m(t).
	\end{equation*}
	This implies that all computations in this paper are carried out entirely within the subspace
	\begin{equation}\label{def:truncated_hilbert_space}
		\widetilde{ \mathcal{H} } := \mathcal{B}^K \times \mathbb{R}^p \in \mathcal{H}.
	\end{equation}
	The $i$th hybrid predictor, projected on $\widetilde{ \mathcal{H} }$, is represented by the tuple
	\begin{equation*}
		\widetilde{W}_i := 
		(
		\widetilde{X}_{i1}, \ldots, \widetilde{X}_{iK}, \boldsymbol{Z}_i
		).
	\end{equation*}
	Let $\boldsymbol{\theta}_{ij}$, $\boldsymbol{\eta}_j$, and $\boldsymbol{\gamma}_j$ denote the $M$-dimensional vectors of coefficients:
	\begin{equation}\label{def:coef_vector_form}
		\boldsymbol{\theta}_{ij} := (\theta_{ij1}, \ldots, \theta_{ijM})^\top,
		\quad
		\boldsymbol{\eta}_{j} := (\eta_{j1}, \ldots, \eta_{jM})^\top ,
		\boldsymbol{\gamma}_{j} := (\gamma_{j1}, \ldots, \gamma_{jM})^\top.
	\end{equation}
	For the predictors, we stack the coefficient vectors across observations into the matrix
	\begin{equation}
		\Theta_j := (\boldsymbol{\theta}_{1j}, \ldots, \boldsymbol{\theta}_{nj})^\top  \in \mathbb{R}^{n \times M}.
	\end{equation}
	The full data for the hybrid PLS problem can be represented by the tuple
	\begin{equation}\label{def:problem_instance}
		(\mathbb{B}, \mathbb{B}^{\prime\prime}, \Theta, \mathbf{y}) \in 
		\mathbb{R}^{(MK+p) \times (MK+p)} \times 
		\mathbb{R}^{(MK+p) \times (MK+p)} \times 
		\mathbb{R}^{n \times (MK+p)} \times 
		\mathbb{R}^n,
	\end{equation}
	where $B, B^{\prime\prime} \in \mathbb{R}^{M \times M}$ are the Gram matrices of the basis functions and their second derivatives, with entries
	\begin{equation}
		B_{m, m'} := \int_0^1 b_m(t)\, b_{m'}(t)\, dt, \quad
		B^{\prime\prime}_{m, m'} := \int_0^1 b_m''(t)\, b_{m'}''(t)\, dt,
	\end{equation}
	for $m, m' = 1, \ldots, M$. We then define the block-diagonal matrices
	\begin{equation}\label{def:gram_block}
		\mathbb{B} := \operatorname{blkdiag}(B, \ldots, B, I_p), \quad
		\mathbb{B}^{\prime\prime} := \operatorname{blkdiag}(B^{\prime\prime}, \ldots, B^{\prime\prime}, I_p),
	\end{equation}
	the full design matrix $\Theta := (\Theta_1, \ldots, \Theta_K, Z) \in \mathbb{R}^{n \times (MK + p)}$, and the response vector $\mathbf{y} := (y_1, \ldots, y_n)^\top$.

@@

	\subsubsection{Proposed Approach: Regularized and Smoothed Computation} \label{sec: Regularization of PLS Components}
Although Proposition \ref{proposition:eigen_noregul} efficiently estimates the PLS component direction $\hat{\xi} = (\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\xi}}_{K+1})$, the functional components $\hat{\xi}_1, \ldots, \hat{\xi}_K$ may not be smooth, potentially leading to overfitting and unstable predictions. To address this, we propose a regularized extension that balances predictive performance with smoothness. Specifically, we penalize the roughness of each $\hat{\xi}_j$ using its integrated squared second derivative $\int_0^1 \bigl\{ \hat{\xi}_j^{\prime\prime}(t) \bigr\}^2 dt$.
Instead of solely maximizing the squared empirical covariance
 $\widehat{\textnormal{Cov}}^2 
\bigl(
\langle \widetilde{W}, \xi \rangle_\mathcal{H}, Y 
\bigr)
$, 
we incorporate this roughness penalty to simultaneously control the complexity of the estimated functional components.

One possible approach is to extend smoothed functional PCA \citep{Rice1991} to our optimization problem \eqref{def:maximizer_squared_empirical_cov}, leading to the following formulation:
	\begin{equation*}
	\hat{\xi} 
	= 
	(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\xi}}_{K+1})
	= \arg \max_{\xi \in \mathcal{B}^K \times \mathbb{R}^p,
		\| \xi  \|_\mathcal{H} = 1
	} 
	\biggl\{
	\widehat{\textnormal{Cov}}^2 
	\bigl(
	\langle \widetilde{W}, \xi \rangle_{\mathcal{H}}, Y 
	\bigr)
	- \sum \limits_{j=1}^K \lambda_j 
	\int_0^1 \bigl\{ \hat{\xi}_j^{\prime\prime}(t) \bigr\}^2 dt
	\biggr\}
	,
\end{equation*}
where the smoothing parameters $\{\lambda_j\}_{j=1}^K$ control the trade-off between maximizing covariance and penalizing roughness. Here, the penalization term can be written in matrix form as:
		\begin{align*}
	\sum \limits_{j=1}^K 
	\lambda_j
	\int_0^1 \bigl\{ \hat{\xi}_j^{\prime\prime}(t) \bigr\}^2 dt
&=
	\sum \limits_{j=1}^K 
\lambda_j
\int_0^1 \bigl\{ \sum_{l=1}^M \theta_{ijl} \, b_l^{\prime \prime}(t) \bigr\}^2 dt
\\&	 =
%%%%%%%%%%%%%%%%%%%
	\sum \limits_{j=1}^K 
\lambda_j
\sum_{l=1}^M
\sum_{m=1}^M
\theta_{ijl} \theta_{ijm}
\int_0^1
 \, b_l^{\prime \prime}(t) 
  b_m^{\prime \prime}(t)
  dt
\\&	 =
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\sum \limits_{j=1}^K
\lambda_j
\boldsymbol{\gamma}_j^\top
B^{\prime \prime} 
\boldsymbol{\gamma}_j
%%%%%%%%%%%%%%%%%%%%%%%%%%
\\&= 
\boldsymbol{\xi}^{\top} \Lambda \mathbb{B}^{\prime \prime} \boldsymbol{\xi},
\end{align*}
where $\boldsymbol{\xi}$ is the concatenated coefficient vector form of $\xi$ as defined in \eqref{def:maximizer_euclidean_form}, $\mathbb{B}^{\prime\prime}$ is defined in \eqref{def:gram_block}, and $\Lambda \in \mathbb{R}^{(MK+p) \times (MK+p)}$ is given by
\begin{equation}\label{def:Lambda}
	\Lambda := \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p}),
\end{equation}
where $0_{p \times p}$ denotes the $p \times p$ zero matrix.

However, the approach of \citet{Rice1991} assumes that the functional predictors admit an orthogonal expansion in the $L^2$ sense. As an alternative, we adopt the strategy of \citet{Silverman1996}, which replaces the standard orthonormality constraint with a weaker one based on a modified inner product that incorporates roughness.
Building on this idea, Proposition \ref{proposition:eigen_regul} modifies the optimization problem presented in Proposition \ref{proposition:eigen_noregul}, yielding a smoothed PLS direction that maximizes the squared covariance over a class of smooth functions.

\begin{proposition}\label{proposition:eigen_regul}
	Let
	$
	(\mathbb{B}, \mathbb{B}^{\prime \prime}, \Theta, \mathbf{y})
	$
	denote the observed data defined in \eqref{def:problem_instance}.
	At the $l$-th iteration of the PLS algorithm, let
	$( \hat{\boldsymbol{\gamma}}_1^\top, \ldots, \hat{\boldsymbol{\gamma}}_K^\top, \hat{\boldsymbol{\xi}}_{K+1}^\top )^\top$
	denote the coefficients of $\hat{\xi}$, defined in \eqref{def:maximizer_squared_empirical_cov},
	which corresponds to the projected unit-norm direction that maximizes the squared empirical covariance.
	These coefficients are obtained by solving the following generalized eigenvalue problem:
	\begin{equation}	\label{eq: Regularized generalized rayleigh quotient equation}
		\left(
		\hat{\boldsymbol{\gamma}}_1^\top, \ldots, \hat{\boldsymbol{\gamma}}_K^\top, \hat{\boldsymbol{\xi}}_{K+1}^\top
		\right)^\top
		=
		\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
		\quad \text{subject to} \quad \boldsymbol{\xi}^\top \mathbb{B} \boldsymbol{\xi} = 1,
	\end{equation}
	where
	\begin{equation}\label{def:V_matrix}
		V := \frac{1}{n^2} (\mathbb{B} \Theta^\top \mathbf{y})(\mathbb{B} \Theta^\top \mathbf{y})^\top \in \mathbb{R}^{(MK+p) \times (MK+p)},
	\end{equation}

\end{proposition}

>>>>

	Given a twice-differentiable basis $\{b_m(t)\}$ of $L^2([0,1])$,  the $j$th functional predictor, regression coefficient, and PLS direction (suppressing iteration indices) are represented as
	\begin{equation*}
		X_{ij}(t) = \sum_{m=1}^\infty \theta_{ijm} \, b_m(t),\quad
		\beta_j(t) = \sum_{m=1}^\infty \eta_{jm} \, b_m(t),\quad
		%%%%%%%%%%%%%%%%%%%%%%
		\xi_j(t)
		=
		\sum_{m=1}^\infty
		\gamma_{jm}
		b_m(t)
	\end{equation*}	
	In practice, the full set of coefficients can not be obtained  with finite sample size, as functional data are measured on a finite grid. Thus, we truncate the expansion at $M$ terms and define the truncated basis space
	$
	\mathcal{B} := \operatorname{span}\bigl(b_1(t), \ldots, b_M(t)\bigr).
	$
	We choose a moderately large $M$ (e.g., 15 or 20) to capture functional variation without fine-tuning, as smoothness is handled via penalization (see Section~\ref{sec: Regularization of PLS Components}).
	The truncated expansions of the predictor and coefficient are denoted as
	\begin{equation*}
		\widetilde{X}_{ij}(t) := \sum_{m=1}^M \theta_{ijm} \, b_m(t), \quad
		\tilde{\beta}_j(t) := \sum_{m=1}^M \eta_{jm} \, b_m(t),
	\end{equation*}
	and our suggest method restricts each PLS direction to admit the following expansion:
	\begin{equation*}
		\xi_j(t) = \sum_{m=1}^M \gamma_{jm} \, b_m(t).
	\end{equation*}
	This implies that all computations in this paper are carried out entirely within the subspace
	\begin{equation}\label{def:truncated_hilbert_space}
		\widetilde{ \mathcal{H} } := \mathcal{B}^K \times \mathbb{R}^p \subset \mathcal{H}.
	\end{equation}
	The $i$th hybrid predictor, projected on $\widetilde{ \mathcal{H} }$, is represented by the tuple
	\begin{equation*}
		\widetilde{W}_i := 
		(
		\widetilde{X}_{i1}, \ldots, \widetilde{X}_{iK}, \boldsymbol{Z}_i
		).
	\end{equation*}
	Let $\boldsymbol{\theta}_{ij}$, $\boldsymbol{\eta}_j$, and $\boldsymbol{\gamma}_j$ denote the $M$-dimensional vectors of coefficients:
	\begin{equation}\label{def:coef_vector_form}
		\boldsymbol{\theta}_{ij} := (\theta_{ij1}, \ldots, \theta_{ijM})^\top,
		\quad
		\boldsymbol{\eta}_{j} := (\eta_{j1}, \ldots, \eta_{jM})^\top ,
		\boldsymbol{\gamma}_{j} := (\gamma_{j1}, \ldots, \gamma_{jM})^\top.
	\end{equation}
	For the predictors, we stack the coefficient vectors across observations into the matrix
	\begin{equation}
		\Theta_j := (\boldsymbol{\theta}_{1j}, \ldots, \boldsymbol{\theta}_{nj})^\top  \in \mathbb{R}^{n \times M}.
	\end{equation}
	The full data for the hybrid PLS problem can be represented by the tuple
	\begin{equation}\label{def:problem_instance}
		(\mathbb{B}, \mathbb{B}^{\prime\prime}, \Theta, \mathbf{y}) \in 
		\mathbb{R}^{(MK+p) \times (MK+p)} \times 
		\mathbb{R}^{(MK+p) \times (MK+p)} \times 
		\mathbb{R}^{n \times (MK+p)} \times 
		\mathbb{R}^n,
	\end{equation}
	where $B, B^{\prime\prime} \in \mathbb{R}^{M \times M}$ are the Gram matrices of the basis functions and their second derivatives, with entries
	\begin{equation}
		B_{m, m'} := \int_0^1 b_m(t)\, b_{m'}(t)\, dt, \quad
		B^{\prime\prime}_{m, m'} := \int_0^1 b_m''(t)\, b_{m'}''(t)\, dt,
	\end{equation}
	for $m, m' = 1, \ldots, M$. We then define the block-diagonal matrices
	\begin{equation}\label{def:gram_block}
		\mathbb{B} := \operatorname{blkdiag}(B, \ldots, B, I_p), \quad
		\mathbb{B}^{\prime\prime} := \operatorname{blkdiag}(B^{\prime\prime}, \ldots, B^{\prime\prime}, I_p),
	\end{equation}
	the full design matrix $\Theta := (\Theta_1, \ldots, \Theta_K, Z) \in \mathbb{R}^{n \times (MK + p)}$, and the response vector $\mathbf{y} := (y_1, \ldots, y_n)^\top$.

@@

	\subsubsection{Proposed Approach: Regularized and Smoothed Computation} \label{sec: Regularization of PLS Components}
Although Proposition \ref{proposition:eigen_noregul} efficiently estimates the PLS component direction $\hat{\xi} = (\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\xi}}_{K+1})$, the functional components $\hat{\xi}_1, \ldots, \hat{\xi}_K$ may not be smooth, potentially leading to overfitting and unstable predictions. To address this, we propose a regularized extension that balances predictive performance with smoothness. Specifically, we penalize the roughness of each $\hat{\xi}_j$ using its integrated squared second derivative 
\begin{equation}
\operatorname{PEN}(\hat{\xi}_j) := \int_0^1 \bigl\{ \hat{\xi}_j^{\prime\prime}(t) \bigr\}^2 dt.
\end{equation}

Instead of solely maximizing the squared empirical covariance
 $\widehat{\textnormal{Cov}}^2 
\bigl(
\langle \widetilde{W}, \xi \rangle_\mathcal{H}, Y 
\bigr)
$, 
we incorporate this roughness penalty to simultaneously control the complexity of the estimated functional components.

One possible approach is to extend the smoothed functional PCA framework of \citet{Rice1991} by modifying the objective in \eqref{def:maximizer_squared_empirical_cov} to
	\begin{equation*}
	\widehat{\textnormal{Cov}}^2 
	\bigl(
	\langle \widetilde{W}, \xi \rangle_{\mathcal{H}}, Y 
	\bigr)
	- \sum \limits_{j=1}^K \lambda_j 
	\operatorname{PEN}({\xi}_j)
	,
\end{equation*}
where the smoothing parameters $\{\lambda_j\}_{j=1}^K$ control the trade-off between maximizing covariance and penalizing roughness. Here, the penalization term can be written in matrix form as:
		\begin{align*}
				\sum \limits_{j=1}^K 
				\lambda_j
				\operatorname{PEN}({\xi}_j)
				&=
	\sum \limits_{j=1}^K 
	\lambda_j
	\int_0^1 \bigl\{ \hat{\xi}_j^{\prime\prime}(t) \bigr\}^2 dt
\\&=
	\sum \limits_{j=1}^K 
\lambda_j
\int_0^1 \bigl\{ \sum_{l=1}^M \theta_{ijl} \, b_l^{\prime \prime}(t) \bigr\}^2 dt
\\&	 =
%%%%%%%%%%%%%%%%%%%
	\sum \limits_{j=1}^K 
\lambda_j
\sum_{l=1}^M
\sum_{m=1}^M
\theta_{ijl} \theta_{ijm}
\int_0^1
 \, b_l^{\prime \prime}(t) 
  b_m^{\prime \prime}(t)
  dt
\\&	 =
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\sum \limits_{j=1}^K
\lambda_j
\boldsymbol{\gamma}_j^\top
B^{\prime \prime} 
\boldsymbol{\gamma}_j
%%%%%%%%%%%%%%%%%%%%%%%%%%
\\&= 
\boldsymbol{\xi}^{\top} \Lambda \mathbb{B}^{\prime \prime} \boldsymbol{\xi},
\end{align*}
where $\boldsymbol{\xi}$ is the concatenated coefficient vector form of $\xi$ as defined in \eqref{def:maximizer_euclidean_form}, $\mathbb{B}^{\prime\prime}$ is defined in \eqref{def:gram_block}, and $\Lambda \in \mathbb{R}^{(MK+p) \times (MK+p)}$ is given by
\begin{equation}\label{def:Lambda}
	\Lambda := \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p}),
\end{equation}
where $0_{p \times p}$ denotes the $p \times p$ zero matrix.

\medskip
\noindent
\textit{Proposed estimation procedure.}~
However, the above approach of \citet{Rice1991} assumes that the functional predictors admit an orthogonal expansion in the $L^2$ sense. As an alternative, we adopt the strategy of \citet{Silverman1996}, which replaces the standard orthonormality constraint with a weaker one based on a modified inner product that incorporates roughness. Accordingly, our estimation procedure at the $l$-th iteration, iteration index omitted and assuming the observations have been residualized in previous steps, solves the following optimization problem:
	\begin{equation}\label{def:maximizer_squared_empirical_cov_reg}
	\hat{\xi} 
	= 
	(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\xi}}_{K+1}
	)
	:= \arg \max_{\xi \in \widetilde{\mathcal{H}}
	}~\widehat{\textnormal{Cov}}^2 
	\bigl(
	\langle \widetilde{W}, \xi \rangle_{\mathcal{H}}, Y 
	\bigr)~s.t.~\| \xi  \|_\mathcal{H} +  \sum \limits_{j=1}^K \lambda_j 
	\operatorname{PEN}(\xi_K) = 1.
\end{equation}
This formulation maximizes the squared covariance over a class of smooth functions.
Proposition~\ref{proposition:eigen_regul} expresses this problem as a generalized Rayleigh quotient.

\begin{proposition}\label{proposition:eigen_regul}
	Let
	$
	(\mathbb{B}, \mathbb{B}^{\prime \prime}, \Theta, \mathbf{y})
	$
	denote the observed data defined in \eqref{def:problem_instance}.
	At the $l$-th iteration of the PLS algorithm, the maximizer of  the squared empirical covariance defined in \eqref{def:maximizer_squared_empirical_cov_reg}, represented as the Euclidean vector form as in \eqref{def:maximizer_euclidean_form},
	is obtained by solving 
	\begin{equation}	\label{eq: Regularized generalized rayleigh quotient equation}
		\left(
		\hat{\boldsymbol{\gamma}}_1^\top, \ldots, \hat{\boldsymbol{\gamma}}_K^\top, \hat{\boldsymbol{\xi}}_{K+1}^\top
		\right)^\top
		=
		\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
		\quad \text{s.t.} \quad \boldsymbol{\xi}^\top (\mathbb{B} + \Lambda \mathbb{B}^{\prime \prime}) \boldsymbol{\xi} = 1.
	\end{equation}
	where we recall that $
		V = n^{-2} (\mathbb{B} \Theta^\top \mathbf{y})(\mathbb{B} \Theta^\top \mathbf{y})^\top$ is defined in  \eqref{def:V_matrix}.
\end{proposition}
